---
title: 'Modelling: LMM'
output:
  html_document: default
  pdf_document: default
date: "`r Sys.Date()`"
---

## Modelling

```{r library and df, results = 'hide'}
setwd("C:/Users/ekaba/0 Uni TUDO/Case Studies/eye_tracking")

library(parameters)
library(dplyr)
library(lme4)
library(ggplot2)
library(clubSandwich)
library(performance)
library(car)
library(colorspace)
library(lmerTest)

df <- read.csv("./results/post_processed_data.csv")

# Reorder Age_Group_Cluster such that "5-18" is the first group
df$Age_Group_Cluster <- factor(df$Age_Group_Cluster,
                               levels = c("5-18", "19-26", "27-35", "36-45", "46-59"))


df$Order_standardized <- (df$Order - mean(df$Order))/sd(df$Order)

#df <- df %>%
# filter(duration < 1.5) #test

# Convert the variable to a factor first
df$ROI <- factor(df$ROI)

# Use relevel to set the reference level to "None"
df$ROI <- relevel(df$ROI, ref = "No_ROI")

df$R_norm <- df$red / 255
df$G_norm <- df$green / 255
df$B_norm <- df$blue / 255

# Create a matrix of normalized RGB values
rgb_matrix <- as.matrix(df[, c('R_norm', 'G_norm', 'B_norm')])

# Convert RGB to HSV
hsv_values <- RGB(rgb_matrix) %>% as("HSV") %>% coords()

# Add HSV values to the data frame
df$Hue <- hsv_values[, 1]
df$Saturation <- hsv_values[, 2]
df$Brightness <- hsv_values[, 3]

# Normalize Hue to [0, 1]
df$Hue_norm <- df$Hue / 360

# Standardize predictors
df$Hue_std <- scale(df$Hue_norm)
df$Saturation_std <- scale(df$Saturation)
df$Brightness_std <- scale(df$Brightness)

# Define color categories based on Hue ranges
df$ColorCategory <- cut(
  df$Hue_norm,
  breaks = c(0, 60/360, 120/360, 180/360, 240/360, 300/360, 1),
  labels = c("Red", "Yellow", "Green", "Cyan", "Blue", "Magenta"),
  include.lowest = TRUE
)

# Convert to factor
df$ColorCategory <- factor(df$ColorCategory)

#df <- df %>% 
#  filter(Valid == 'True')

```


```{r functions, echo=FALSE}
plot_residuals <- function(model){
  
  residuals_model <- residuals(model, type = "pearson")
  
  # Plot residuals vs fitted values
  fitted_values <- fitted(model)
  plot(fitted_values, residuals_model, xlab = "Fitted Values", ylab = "Residuals",
       main = "Residuals vs Fitted Values")
  abline(h = 0, col = "red", lty = 2)
  
  # Q-Q plot of residuals
  qqnorm(residuals_model, main = "Q-Q Plot of Residuals")
  qqline(residuals_model, col = "red")
  
  #create histogram of residuals
  hist(residuals_model, xlab = "Residuals", main = "Histogram of the residuals")
  
  predicted_values <- predict(model)
  
  # Create a data frame with actual and predicted values
  plot_data <- data.frame(
    Actual_Log_Duration = df$log_duration,  # Actual log-transformed duration
    Predicted_Log_Duration = predicted_values
  )
  
  min_value <- min(plot_data$Predicted_Log_Duration, plot_data$Actual_Log_Duration)
  max_value <- max(plot_data$Predicted_Log_Duration, plot_data$Actual_Log_Duration)
  
  # Add a trend line to see if there are systematic deviations
  ggplot(plot_data, aes(x = Predicted_Log_Duration, y = Actual_Log_Duration)) +
    geom_point(alpha = 0.5) +
    geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
    geom_smooth(method = "lm", color = "blue") +  # Add a smooth trend line
    labs(
      title = "Actual vs. Predicted Log Duration with Trend Line",
      x = "Predicted Log Duration",
      y = "Actual Log Duration"
    ) +
    theme_minimal() +
    coord_fixed(ratio = 1) + # Set the aspect ratio to 1:1
    xlim(min_value, max_value) +  # Set x-axis limits from min to max
    ylim(min_value, max_value)    # Set y-axis limits from min to max
} 

get_robust_se <- function(model, df){
  
  robust_se <- vcovCR(model, cluster = df$ID, type = "CR2")
  summary_robust <- coef_test(model, vcov = robust_se)
  
  print(summary_robust)
}


random_effects_diagnostics <- function(model, effect_name = "a") {
  # Extract all random effects
  random_effects <- ranef(model)

#Check for the random effect column name
  if (!effect_name %in% colnames(random_effects)) {
    stop(paste("Effect", effect_name, "not found. Available effects are:",
               paste(colnames(random_effects), collapse = ", ")))
  }

#Extract the specified random effect
  effect_values <- random_effects[, effect_name]

#Q-Q plot for the random effect
  qqnorm(effect_values, 
         main = paste("Q-Q Plot for Random Effect:", effect_name), 
         pch = 20, col = "blue")
  qqline(effect_values, col = "red", lty = 2)

#Return the random effect values for further use
  return(effect_values)
}

```

We use standardized order since, when using squared order later as specification, the number of order^2 can get very big.

```{r models}
# base model
model1<- lmer(log_duration ~ Order_standardized + (1 | ID), data = df)
print(r2(model1))
model2 <- lmer(log_duration ~ Order_standardized  + I(Order_standardized^2) + (1 | ID), data = df)
print(r2(model2))
anova(model1, model2)


model3 <- lmer(log_duration ~ Order_standardized + I(Order_standardized^2) + ColorCategory + Saturation_std + Brightness_std + (1 | ID), data = df)
print(r2(model3))
anova(model2, model3)

model4 <- lmer(log_duration ~ Order_standardized  + I(Order_standardized^2) + ColorCategory + Saturation_std + Brightness_std + ROI + (1 | ID), data = df)
anova(model3, model4)
print(r2(model4))

model5 <- lmer(log_duration ~ Order_standardized  + I(Order_standardized^2) + ColorCategory + Saturation_std + Brightness_std + ROI + Gender + Age_Group_Cluster + (1 | ID), data = df)
print(r2(model5))
anova(model4, model5)

get_robust_se(model5, df)
```

Order has a significant positive effect on log duration. This result also holds if we use standard errors that are robust against heteroscedasticity. Also, the SD of the random intercept for ID (0.12) suggests that the random effect of each participant explains a moderate amount of the variation in log duration, but there is still a lot unexplained by the model.

R2m (marginal r-squared), is the effect size without considering the random effects (i.e., where the model is built without regard to participant differences). The second, R^2c (conditional r-squared) is the effect size when the random effects are taken into account. Accordingly, the R^2c values are aways higher than the R^2m values (unless there are little/no random effects).

Random effects explain a lot more of the variance in log duration than without. But it is still not a lot, with only about 4.7%.

```{r final model plot residuals}
plot_residuals(model5)
vif(model5)
```

Arguably roughly normal, but the residuals show signs of heteroscedasticity and also actual vs predicted suggests that linearity assumption might not hold well. This model kinda predicts log durations around the mean.

Now we will increase complexity of the model and look if it significantly improves the fit. Since we have concerns about heteroscedasticity, we will also use robust estimates.

To compare the fits of two models, you can use the anova() function with the regression objects as two separate arguments. The anova() function will take the model objects as arguments, and return an ANOVA testing whether the more complex model is significantly better at capturing the data than the simpler model. If the resulting p-value is sufficiently low (usually less than 0.05), we conclude that the more complex model is significantly better than the simpler model, and thus favor the more complex model. If the p-value is not sufficiently low (usually greater than 0.05), we should favor the simpler model.

source: https://bookdown.org/ndphillips/YaRrr/comparing-regression-models-with-anova.html

Order^2 has a significant negative effect on duration, i.e., order first increases duration but eventually decreases it. Also, the model with order^2 has a significantly better fit than without.
The added Order^2 is able to explain more than the base model (higher R^2m), but the R^2c is a little less than before. Probably because the effect of Order^2 was attributed to random effects before.

Some ROIs are significant, and the ANOVA test suggest that including ROI leads to a better fit. ROI area not significant.

Also adding HSV values lead to a better fit than without.

Adding HSV values leads to better fit than without compared to ROI model. 

```{r normality of random effects}
library(checkmate)
library(cowplot)
library(ggplot2)
library(purrr)
library(tidyr)
library(qqplotr)

plot_ranef <- function(model){

  # Return an error if an acceptable model type is not entered in the function
  checkmate::expect_class(model, "lmerMod",
                          info = "The input model is not accepted by plot_raneff. Model must be fit using 'lmer'.")

  # building dataframe of all random effects
  bmat <- as.data.frame(lme4::ranef(model))

  # converting each random effect vector into one line with nest
  renest <- tidyr::nest(bmat, data = c("grp", "condval", "condsd"))

  # generating list of ggplot objects for each random effect vector
  plots <- purrr::pmap(list(renest$data, renest$grpvar, renest$term),
                          function(a,b,c){
                            ggplot(data = a, aes_string(sample = "condval")) +
                              qqplotr::stat_qq_band(bandType = "pointwise",
                                                    distribution = "norm",
                                                    fill = "#FBB4AE", alpha = 0.4) +
                              qqplotr::stat_qq_line(distribution = "norm", colour = "#FBB4AE") +
                              qqplotr::stat_qq_point(distribution = "norm") +
                              xlab("Normal quantiles") + theme_bw() +
                              ylab(paste(b,": ", c)) +
                              ggtitle('Q-Q plot of the random effect') + 
                              theme(plot.title = element_text(hjust = 0.5))
                          }
  )

  # Create grid of all random effect plots
  cowplot::plot_grid(plotlist = plots)
}

plot_ranef(model5)
```
Code from https://rdrr.io/github/goodekat/redres/src/R/plot_ranef.R

## Looking at specific age clusters

```{r other models 3}
# Split the dataset by age group 
df$age_std <- scale(df$age)

df_5_18 <- subset(df, Age_Group_Cluster == "5-18")
df_19_26 <- subset(df, Age_Group_Cluster == "19-26")
df_27_35 <- subset(df, Age_Group_Cluster == "27-35")
df_36_45 <- subset(df, Age_Group_Cluster == "36-45")
df_46_59 <- subset(df, Age_Group_Cluster == "46-59")

# Age Group 5-18
model_5_18 <- lmer(log_duration ~ Order_standardized + I(Order_standardized^2) + age_std +  Gender  + ROI + ColorCategory + Saturation_std + Brightness_std + (1 | ID), data = df_5_18)
# Age Group 19-26
model_19_26 <- lmer(log_duration ~ Order_standardized + I(Order_standardized^2) + age_std + Gender + ROI + ColorCategory + Saturation_std + Brightness_std  + (1 | ID), data = df_19_26)
# Age Group 27-35
model_27_35 <- lmer(log_duration ~ Order_standardized + I(Order_standardized^2) + age_std + Gender + ROI + ColorCategory + Saturation_std + Brightness_std + (1 | ID), data = df_27_35)
# Age Group 36-45
model_36_45 <- lmer(log_duration ~ Order_standardized + I(Order_standardized^2) + age_std + Gender  + ROI + ColorCategory + Saturation_std + Brightness_std  + (1 | ID), data = df_36_45)
# Age Group 46-59
model_46_59 <- lmer(log_duration ~ Order_standardized + I(Order_standardized^2) + age_std + Gender  + ROI + ColorCategory + Saturation_std + Brightness_std  + (1 | ID), data = df_46_59)

get_robust_se(model_5_18, df_5_18)
get_robust_se(model_19_26, df_19_26)
get_robust_se(model_27_35, df_27_35)
get_robust_se(model_36_45, df_36_45)
get_robust_se(model_46_59, df_46_59)

print(r2(model_5_18))
print(r2(model_19_26))
print(r2(model_27_35))
print(r2(model_36_45))
print(r2(model_46_59))


```


For the age group 5-18, there is a significant (10%) negative effect of age on log duration. Also for the other age groups, age is not significant. Age group 18-26 is the only group, where GenderMALE is significant (5%), having a lower mean log duration than females.

```{r plots, echo=FALSE}

# Plot for the youngest age group (5-18)
ggplot(df_5_18, aes(x = age, y = log_duration)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "darkblue", se = FALSE) +
  labs(title = "Age vs Log Fixation Duration (Age Group 5-18)",
       x = "Age",
       y = "Log Fixation Duration") +
  theme_minimal()

# Plot for the oldest age group (46-59)
ggplot(df_46_59, aes(x = age, y = log_duration)) +
  geom_point(alpha = 0.5, color = "red") +
  geom_smooth(method = "lm", color = "darkred", se = FALSE) +
  labs(title = "Age vs Log Fixation Duration (Age Group 46-59)",
       x = "Age",
       y = "Log Fixation Duration") +
  theme_minimal()

# Combine both datasets
df_combined <- rbind(df_5_18, df_46_59)

```

## Summary:
Order is significantly positive and order^2 significantly negative: Fixations become longer at first but then at a certain point (fixation order) become shorter again.

The age clusters were not significantly affecting the duration in the full model
Looking at the individual age clusters, only for 5-18 age was 'significant' predictor, showing a decrease in fixation duration with increasing age. But weak significance.

Gender was mostly insignificant, only for the cluster 18-26, where males have a shorter average log duration fixation.

BUT: Model seems to explain very little of the total variation in log fixation duration. We miss key predictors, order, age and gender, ROIs, HSB alone are not enough. 



