---
title: "nlme_final"
author: "Carter Hogan"
date: "11/22/2024"
output: html_document
---

```{r setup, include=FALSE}
# attach relevant libraries
library(nlme)
library(dplyr)
library(sandwich)
# read in csv of data
# set working directory
setwd('C:/Users/chica/OneDrive/Documents/Case Studies/eye_tracking/results')
# read in the data
df <- read.csv('post_processed_data.csv')
# create a log variable for Order
df$Order_log <- log(df$Order + 1)
df$ROI_area <- ifelse(is.na(df$ROI_area), 0, df$ROI_area)
df$ROI_area <- scale(ifelse(is.na(df$ROI_area), 0, df$ROI_area))
df$ROI <- ifelse(is.na(df$ROI), "No_ROI", df$ROI)
df$ROI <- factor(df$ROI)
df$ROI <- factor(df$ROI, levels =c("No_ROI","airship",
                                   "coastal_town","diver_with_abs",
                                   "elefant_under_water","female_diver",
                                   "island_with_lighttower",
                                   "little_boat_in_the_distance", 
                                   "man_wakeboarding","mermaid", "octupus",
                                   "people_jumping_from_boat","seal",         
                                   "son_and_father_fishing","space",
                                   "sunset","surfboard","turtle",
                                   "woman_on_surboard"))
```

##Non-Linear Mixed Effects Models

#Model Specification



```{r simple}
model_simple <- nlme(
  log_duration ~ a - b * Order_log,
  fixed = a + b ~1,            # Intercept-only fixed effects
  random = a + b ~ 1 | ID,
  start = c(a_intercept = 100,            # Initial value for a's intercept
            b_intercept = 10),
  data = df
)

summary(model_simple)
```
```{r Age}
model_age<- nlme(
  log_duration ~ a - b * Order_log,
  fixed = a + b ~ Age_19.26 + Age_27.35 + Age_36.45 + Age_46.59,            # Intercept-only fixed effects
  random = a + b ~ 1 | ID,
  start = c(a_intercept = -1.642777,            # Initial value for a's intercept
            b_intercept = -0.04070717,
            a_19_26 = 0,                    
            b_19_26 = 0,                   
            a_27_35 = 0,                
            b_27_35 = 0,                   
            a_36_45 = 0,                
            b_36_45 = 0,                
            a_46_59 = 0,                
            b_46_59 = 0
            ),
  data = df
)


```

# Shows that age group clusters improves the AIC


```{r age and gender, echo=FALSE}
model_age_gender <- nlme(
  duration ~ a - b * Order_log,
  fixed = a + b ~ Age_19.26 + Age_27.35 + Age_36.45 + Age_46.59 +female,            # Intercept-only fixed effects
  random = a + b ~ 1 | ID,
  start = c(a_intercept = -1.642777,            # Initial value for a's intercept
            b_intercept = -0.04070717,
            a_19_26 = 0,                    
            b_19_26 = 0,                   
            a_27_35 = 0,                
            b_27_35 = 0,                   
            a_36_45 = 0,                
            b_36_45 = 0,                
            a_46_59 = 0,                
            b_46_59 = 0, 
            a_female = 0,
            b_female = 0
            
           ),
  control = nlmeControl(msMaxIter = 300),
  data = df
)

```

```{r ROI}

model_roi_cat <- nlme(
  log_duration ~ a - b * Order_log,
  fixed = a + b ~ Age_19.26 + Age_27.35 + Age_36.45 + Age_46.59 + female + ROI,
  random = a + b ~ 1 | ID,
  start = c(a_intercept = -1.642777,            # Initial value for a's intercept
            b_intercept = -0.04070717,
            a_19_26 = 0,                    
            b_19_26 = 0,                   
            a_27_35 = 0,                
            b_27_35 = 0,                   
            a_36_45 = 0,                
            b_36_45 = 0,                
            a_46_59 = 0,                
            b_46_59 = 0, 
            a_female = 0,
            b_female = 0,
            a_airship = 0,
            b_airship = 0,a_coastal_town = 0,
            b_coastal_town = 0,a_diver = 0,
            b_diver = 0,a_elef = 0,
            b_elef = 0,a_f_diver = 0,
            b_f_diver = 0,a_isl = 0,
            b_isl = 0,a_boat = 0,
            b_boat = 0,a_wake = 0,
            b_wake = 0,a_mermaid = 0,
            b_mermaid = 0,a_oct = 0,
            b_oct = 0,a_jump = 0,
            b_jump = 0,a_seal = 0,
            b_seal = 0,a_son = 0,
            b_son = 0,a_space = 0,
            b_space = 0,a_sunset = 0,
            b_sunset = 0,a_surf = 0,
            b_surf = 0,a_turtle = 0,
            b_turtle = 0,a_woman_surf = 0,
            b_woman_surf = 0),
  control = nlmeControl(maxIter = 200 ),
  data = df
)
```
# It shows me that the AIC is worse for ROI_cat

```{r r_squared}
r2_nlme <- function(model) {
fixed_effects_predictions <- predict(model, level = 0)
var_fixed <- var(fixed_effects_predictions)
var_random <- as.numeric(VarCorr(model)[1, "Variance"])  # Random effects variance
var_residual <- model$sigma^2                             # Residual variance
r2_marginal <- var_fixed / (var_fixed + var_random + var_residual)
r2_conditional <- (var_fixed + var_random) / (var_fixed + var_random + var_residual)
  
  # Return results
  return(list(
    Marginal_R2 = r2_marginal,
    Conditional_R2 = r2_conditional,
    Residual_Variance = var_residual,
    Random_Effects_Variance = var_random,
    Fixed_Effects_Variance = var_fixed
  ))
}

```

```{r}
df$Age_19 <- df$Age_19.26
df$Age_27 <- df$Age_27.35
df$Age_36 <- df$Age_36.45
df$Age_46 <- df$Age_46.59
df$bright <- df$Brightness_normed
df$sat <- df$Saturation_normed
```

```{r sat}
model_sat_bright <-  nlme(
  log_duration ~ a - b * Order_log,
  fixed = a + b ~ bright + sat + Age_19 + Age_27 + Age_36 + Age_46 + female + ROI,
  random = a + b ~ 1 | ID,
  start = c(a_intercept = -1.642777,            # Initial value for a's intercept
            b_intercept = -0.04070717, a_bright = 0, b_bright= 0, a_sat = 0, b_sat = 0,
            a_19_26 = 0,                    
            b_19_26 = 0,                   
            a_27_35 = 0,                
            b_27_35 = 0,                   
            a_36_45 = 0,                
            b_36_45 = 0,                
            a_46_59 = 0,                
            b_46_59 = 0, 
            a_female = 0,
            b_female = 0,
            a_airship = 0,
            b_airship = 0,a_coastal_town = 0,
            b_coastal_town = 0,a_diver = 0,
            b_diver = 0,a_elef = 0,
            b_elef = 0,a_f_diver = 0,
            b_f_diver = 0,a_isl = 0,
            b_isl = 0,a_boat = 0,
            b_boat = 0,a_wake = 0,
            b_wake = 0,a_mermaid = 0,
            b_mermaid = 0,a_oct = 0,
            b_oct = 0,a_jump = 0,
            b_jump = 0,a_seal = 0,
            b_seal = 0,a_son = 0,
            b_son = 0,a_space = 0,
            b_space = 0,a_sunset = 0,
            b_sunset = 0,a_surf = 0,
            b_surf = 0,a_turtle = 0,
            b_turtle = 0,a_woman_surf = 0,
            b_woman_surf = 0),
  control = nlmeControl(maxIter = 200 ),
  data = df
)



```
```{r}
#simple
m1<-r2_nlme(model_simple)
#age
m2<-r2_nlme(model_age)
# age + gender
m3<-r2_nlme(model_age_gender)
# age + gender + roi_cat
m4<-r2_nlme(model_roi_cat)
## age + gender + roi_cat + bright + sat
m5<-r2_nlme(model_sat_bright)


model_summary <- data.frame(
  Model = c("Simple", "Age", "Age + Gender", "Age + Gender + ROI_Cat", "Age + Gender + ROI_Cat + Bright + Sat"), 
  Marginal_R2 = c(m1$Marginal_R2, m2$Marginal_R2, m3$Marginal_R2, 
                  m4$Marginal_R2, m5$Marginal_R2),
  Conditional_R2 = c(m1$Conditional_R2, m2$Conditional_R2, 
                     m3$Conditional_R2, m4$Conditional_R2, m5$Conditional_R2),
  Residual_Variance = c(m1$Residual_Variance, m2$Residual_Variance, 
                        m3$Residual_Variance, m4$Residual_Variance,
                        m5$Residual_Variance),
  Random_Effects_Variance = c(m1$Random_Effects_Variance, 
                              m2$Random_Effects_Variance, 
                              m3$Random_Effects_Variance, 
                              m4$Random_Effects_Variance, 
                              m5$Random_Effects_Variance),
  Fixed_Effects_Variance = c(m1$Fixed_Effects_Variance, m2$Fixed_Effects_Variance, m3$Fixed_Effects_Variance, m4$Fixed_Effects_Variance, m5$Fixed_Effects_Variance)
)

```
# We see that Model 5 performs best in terms of marginal R^2 and conditional R^2, models 2 and 3 have lower residual vairance indicating tighter fits, the random effects variance is roughly the same in each model at ~0.34 indicating the random effects are useful in explaining the variation
```{r normality function}
residual_diagnostics <- function(model) {
  # Extract residuals
  res <- residuals(model)
  fitted_vals <- fitted(model)
  
  # Set up plotting area
  par(mfrow = c(2, 2))  # 2x2 grid for plots
  
  # Residuals vs Fitted Values
  plot(fitted_vals, res, 
       xlab = "Fitted Values", 
       ylab = "Residuals", 
       main = "Residuals vs Fitted Values",
       pch = 20, col = "blue")
  abline(h = 0, col = "red", lty = 2)
  
  # Q-Q Plot
  qqnorm(res, main = "Q-Q Plot of Residuals", pch = 20, col = "blue")
  qqline(res, col = "red", lty = 2)
  
  # Histogram of Residuals
  hist(res, 
       breaks = 20, 
       main = "Histogram of Residuals", 
       xlab = "Residuals", 
       col = "lightblue", 
       border = "black")


}
```

```{r Normality}
residual_diagnostics(model_simple)
#age
residual_diagnostics(model_age)
# age + gender
residual_diagnostics(model_age_gender)
# age + gender + roi_cat
residual_diagnostics(model_roi_cat)
## age + gender + roi_cat + bright + sat
residual_diagnostics(model_sat_bright)

```
```{r Random Effects Function}
random_effects_diagnostics <- function(model, effect_name = "a") {
  # Extract all random effects
  random_effects <- ranef(model)
  
  # Check for the random effect column name
  if (!effect_name %in% colnames(random_effects)) {
    stop(paste("Effect", effect_name, "not found. Available effects are:",
               paste(colnames(random_effects), collapse = ", ")))
  }
  
  # Extract the specified random effect
  effect_values <- random_effects[, effect_name]
  
  # Q-Q plot for the random effect
  qqnorm(effect_values, 
         main = paste("Q-Q Plot for Random Effect:", effect_name), 
         pch = 20, col = "blue")
  qqline(effect_values, col = "red", lty = 2)
  
  # Return the random effect values for further use
  return(effect_values)
}
```

```{r Random Effects Assumptions}
random_effects_diagnostics(model_simple,effect_name = "a")
#age
random_effects_diagnostics(model_age,effect_name = "a.(Intercept)")
# age + gender
random_effects_diagnostics(model_age_gender,effect_name = "a.(Intercept)")
# age + gender + roi_cat
random_effects_diagnostics(model_roi_cat,effect_name = "a.(Intercept)")
## age + gender + roi_cat + bright + sat
random_effects_diagnostics(model_sat_bright,effect_name = "a.(Intercept)")
```
```{r aic_bic func}
# Function to calculate AIC and BIC for a list of models
get_aic_bic <- function(models) {
  # Initialize empty vectors to store AIC and BIC values
  aic_values <- numeric(length(models))
  bic_values <- numeric(length(models))
  
  # Loop through each model in the list
  for(i in seq_along(models)) {
    # Calculate AIC and BIC for the current model
    aic_values[i] <- AIC(models[[i]])
    bic_values[i] <- BIC(models[[i]])
  }
  
  # Create a data frame with model names, AIC, and BIC
  model_names <- names(models)
  results <- data.frame(
    Model = model_names,
    AIC = aic_values,
    BIC = bic_values
  )
  
  return(results)
}


```
```{r}

models_list <- list(
  model_simple = model_simple,
  model_age = model_age,
  model_age_gender = model_age_gender,
  model_roi_cat = model_roi_cat,
  model_sat_bright = model_sat_bright
)

# Get AIC and BIC for all models
model_comparison <- get_aic_bic(models_list)
print(model_comparison)
```


```{r Diagnostics}
residual_diagnostics(model_sat_bright)
random_effects_diagnostics(model_sat_bright,effect_name = "a.(Intercept)")

models_list <- list(
  model_simple = model_simple,
  model_age = model_age,
  model_age_gender = model_age_gender,
  model_roi_cat = model_roi_cat,
  model_sat_bright = model_sat_bright,
  model_final = model_final
)

# Get AIC and BIC for all models
model_comparison <- get_aic_bic(models_list)
print(model_comparison)


m6<-r2_nlme(model_final)
# Create a new row for the sixth model
new_row <- data.frame(
  Model = "Sixth Model", 
  Marginal_R2 = m6$Marginal_R2, 
  Conditional_R2 = m6$Conditional_R2,
  Residual_Variance = m6$Residual_Variance,
  Random_Effects_Variance = m6$Random_Effects_Variance,
  Fixed_Effects_Variance = m6$Fixed_Effects_Variance
)

# Append the new row to the existing model_summary
model_summary <- rbind(model_summary, new_row)
```




```{r}
# Get the model summary
summary_model <- summary(model_sat_bright)

# Extract coefficients (fixed effects)
coef_summary <- summary_model$tTable

# Create a nice table for display
coef_table <- data.frame(
  "Parameter" = rownames(coef_summary),
  "Estimate" = coef_summary[, "Value"],
  "Std. Error" = coef_summary[, "Std.Error"],
  "t-value" = coef_summary[, "t-value"],
  "p-value" = coef_summary[, "p-value"]
)

```




```{r}

coef_summary <- coef_table[coef_table$p.value <0.05,]

```


```{r}
plot(fitted(model_sat_bright), residuals(model_sat_bright),
     xlab = "Fitted Values", ylab = "Residuals", 
     main = "Residuals vs Fitted Values")
abline(h = 0, col = "red")

# Normal Q-Q plot of residuals
qqnorm(residuals(model_sat_bright))
qqline(residuals(model_sat_bright), col = "red")
```
```{r}
# Extract fixed effects (coefficients)
coefficients <- fixef(model_sat_bright)

# Calculate the robust variance-covariance matrix using the sandwich estimator
vcov_robust <- vcov(model_sat_bright, type = "HC3")  # Robust standard errors using HC3

# Extract robust standard errors (diagonal elements of the covariance matrix)
robust_se <- sqrt(diag(vcov_robust))

# Calculate z-values (coefficient / robust SE)
z_values <- coefficients / robust_se

# Calculate robust p-values (2-tailed)
p_values_robust <- 2 * pnorm(-abs(z_values))  # Two-tailed test

# Create a summary table with coefficient names, robust standard errors, and p-values
summary_table <- data.frame(
  Coefficient = names(coefficients),
  Estimate = coefficients,
  Robust_SE = robust_se,
  Z_value = z_values,
  P_value = p_values_robust
)

# Filter for significant coefficients (p-value < 0.05)
significant_summary_table <- summary_table[summary_table$P_value < 0.05, ]

# Print the summary table
print(significant_summary_table)

  
```
```{r}
# Add stars based on p-value significance
significant_summary_table$Significance <- ifelse(significant_summary_table$P_value < 0.001, "***",
                                       ifelse(significant_summary_table$P_value < 0.01, "**",
                                       ifelse(significant_summary_table$P_value < 0.05, "*", "")))

# Create LaTeX table with Coefficients, Estimates, P-values, and Significance stars
latex_table <- paste0(
  "\\begin{table}[ht]\n",
  "\\centering\n",
  "\\begin{tabular}{|l|c|c|c|}\n",
  "\\hline\n",
  "Coefficient & Estimate & P-value & Significance \\\\ \n",
  "\\hline\n"
)

# Add the significant rows to the LaTeX table
for (i in 1:nrow(significant_summary_table)) {
  latex_table <- paste0(
    latex_table,
    significant_summary_table$Coefficient[i], " & ",
    round(significant_summary_table$Estimate[i], 4), " & ",
    round(significant_summary_table$P_value[i], 4), " & ",
    significant_summary_table$Significance[i], " \\\\ \n"
  )
}

# Close the LaTeX table
latex_table <- paste0(latex_table, "\\hline\n", "\\end{tabular}\n", "\\end{table}")

# Print the LaTeX table
cat(latex_table)
```
```{r model_summary}
library(xtable)
# Convert to LaTeX table
latex_table <- xtable(model_summary,digits = c(0, 0, 4, 4, 4, 4, 4))

# Print the LaTeX code
print(latex_table, type = "latex", include.rownames = FALSE)
```
```{r model_comparison}

# Create LaTeX table with custom decimal places
latex_table <- xtable(model_comparison)

# Print the LaTeX code
print(latex_table, type = "latex", include.rownames = FALSE)
```
```{r}
png("nlme_diagnostics.png", width = 12, height = 6, units = "in", res = 300)
residual_diagnostics(model_sat_bright)


```
```{r}
png("random_effects_diagnostics_wide.png", width = 12, height = 6, units = "in", res = 300)

# Set up a 1x2 plot layout (side-by-side)
par(mfrow = c(1, 2)) 
random_effects_diagnostics(model_sat_bright,effect_name = "a.(Intercept)")
random_effects_diagnostics(model_sat_bright,effect_name = "b.(Intercept)")
```

```{r}
#Add stars based on p-value significance
summary_table$Significance <- ifelse(summary_table$P_value < 0.001, "***",
                                       ifelse(summary_table$P_value < 0.01, "**",
                                       ifelse(summary_table$P_value < 0.05, "*", "")))

# Create LaTeX table with Coefficients, Estimates, P-values, and Significance stars
latex_table <- paste0(
  "\\begin{table}[ht]\n",
  "\\centering\n",
  "\\begin{tabular}{|l|c|c|c|}\n",
  "\\hline\n",
  "Coefficient & Estimate & P-value & Significance \\\\ \n",
  "\\hline\n"
)

# Add the significant rows to the LaTeX table
for (i in 1:nrow(significant_summary_table)) {
  latex_table <- paste0(
    latex_table,
    summary_table$Coefficient[i], " & ",
    round(summary_table$Estimate[i], 4), " & ",
    round(summary_table$P_value[i], 4), " & ",
    summary_table$Significance[i], " \\\\ \n"
  )
}

# Close the LaTeX table
latex_table <- paste0(latex_table, "\\hline\n", "\\end{tabular}\n", "\\end{table}")

# Print the LaTeX table
cat(latex_table)
```
```{r}
library(xtable)
# Get the summary of the model
model_summary <- summary(model_sat_bright)

# Extract coefficients and p-values
coefficients <- as.data.frame(model_summary$tTable)

# Add row names as a column for better clarity
coefficients$Parameter <- rownames(coefficients)

# Rearrange columns to show parameter, estimate, and p-value
coefficients <- coefficients[, c("Parameter", "Value", "p-value")]

# Convert to a LaTeX table using xtable
latex_table <- xtable(coefficients, caption = "Nonlinear Mixed-Effects Model Coefficients")

# Print the LaTeX table in the sidewaystable environment
print(
  latex_table, 
  include.rownames = FALSE,
  floating.environment = "sidewaystable", # Use sidewaystable
  table.placement = "ht" # Optional placement
)

```

